---
title: "Practice Lectures Week 8 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "Mar 28, 2022"
---

# Nonlinear Linear Regression: Polynomial Regression

Let's generate a polynomial model of degree 2 ($f = h^* \in \mathcal{H}$) and let $\epsilon$ be random noise (the error due to ignorance) for $\mathbb{D}$ featuring $n = 2$.

```{r}
set.seed(1003)
n = 25
beta_0 = 1
beta_1 = 0
beta_2 = 1
x = runif(n, -2, 5)
#best possible model
h_star_x = beta_0 + beta_1 * x + beta_2 * x^2

#actual data differs due to information we don't have
epsilon = rnorm(n)
y = h_star_x + epsilon

#scatterplot it
df = data.frame(x = x, y = y, h_star_x = h_star_x)
pacman::p_load(ggplot2)
basic = ggplot(df, aes(x, y)) +
  geom_point()
basic
head(x)
```

Let's try to estimate with a line:

```{r}
linear_mod = lm(y ~ x)
b_linear = summary(linear_mod)$coef
basic + geom_abline(intercept = b_linear[1], slope = b_linear[2], col = "red")
```

The relationship is "underfit". $\mathcal{H}$ is not rich enough right now to express something close to $f(x)$. But it is better than the null model!

Now let's do a polynomial regression of degree two. Let's do so manually:

```{r}
X = as.matrix(cbind(1, x, x^2))
head(X)
b = solve(t(X) %*% X) %*% t(X) %*% y
b
c(beta_0, beta_1, beta_2)
```

These are about the same as the $\beta_0, \beta_1$ and $\beta_2$ as defined in $f(x)$ the true model. In order to graph this, we can no longer use the routine `geom_abline`, we need to use `stat_function`.

```{r}
plot_function_degree_2 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2
}

basic + 
  stat_function(fun = plot_function_degree_2, args = list(b = b), col= "red") + 
  stat_function(fun = plot_function_degree_2, args = list(b = c(beta_0, beta_1, beta_2)), col= "darkgreen")
```

Now let's try polynomial of degree 3:

```{r}
X = as.matrix(cbind(1, x, x^2, x^3))
b = solve(t(X) %*% X) %*% t(X) %*% y
b

plot_function_degree_3 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3
}

basic + stat_function(fun = plot_function_degree_3, args = list(b = b), col = "red") + 
  stat_function(fun = plot_function_degree_2, args = list(b = c(beta_0, beta_1, beta_2, 0)), col= "darkgreen")
```

Still the same. Why? The $x^3$ term is like adding one "nonsense" predictor. One nonsense predictor marginally affects $R^2$ but it doesn't matter too much.

Now let's try polynomial of degree 8:

```{r}
X = as.matrix(cbind(1, x, x^2, x^3, x^4, x^5, x^6, x^7, x^8))
b = solve(t(X) %*% X) %*% t(X) %*% y
b

plot_function_degree_8 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4 + b[6] * x^5 + b[7] * x^6 + b[8] * x^7 + b[9] * x^8 
}

basic + stat_function(fun = plot_function_degree_8, args = list(b = b), col = "red") + 
  stat_function(fun = plot_function_degree_2, args = list(b = c(beta_0, beta_1, beta_2, rep(0, 6))), col= "darkgreen")
```

We are seeing now a little bit of "overfitting" in the edge(s). We now have $p=9$ and $n=100$. We can do a lot worse!

Let's learn how to do this in R first without having to resort to manual linear algebra. R has a function called "poly" that can be used *inside* formula declarations.

Let's first fit the degree 2 model:

```{r}
degree_2_poly_mod = lm(y ~ poly(x, 2, raw = TRUE))
head(model.matrix(~ poly(x, 2, raw = TRUE))) #the model matrix for this regression - just to check
b_poly_2 = coef(degree_2_poly_mod)
b_poly_2
summary(degree_2_poly_mod)$r.squared
```

Let's go on a slight tangent. And look at this regression without using the raw polynomial.

```{r}
Xmm = model.matrix(~ poly(x, 2))
head(Xmm) #the model matrix for this regression - just to check
Xmm[, 1] %*% Xmm[, 2]
Xmm[, 2] %*% Xmm[, 2]
Xmm[, 2] %*% Xmm[, 3]
Xmm[, 3] %*% Xmm[, 3]
```

Are thse orthogonal polynomials? How is the `poly` function without `raw = TRUE` working to generate a model matrix?

```{r}
degree_2_orthog_poly_mod = lm(y ~ poly(x, 2))
b_poly_2 = coef(degree_2_orthog_poly_mod)
b_poly_2
summary(degree_2_orthog_poly_mod)$r.squared
```

Raw or orthogonal does not affect the yhats and Rsq. They are the same as we got before! That's because the colspace is the same in both cases raw or polynomial. We use "raw" polynomials to keep them interpretable and on the same scale as the manual models we were fitting.

Now let's do polynomial of degree 13:

```{r}
degree_13_poly_mod = lm(y ~ poly(x, 13, raw = TRUE))
b_poly_13 = coef(degree_13_poly_mod)

plot_function_degree_13 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4 + b[6] * x^5 + b[7] * x^6 + b[8] * x^7 + b[9] * x^8 + b[10] * x^9  + b[11] * x^10 + b[12] * x^11 + b[13] * x^12 + b[14] * x^13
}

basic + stat_function(fun = plot_function_degree_13, args = list(b = b_poly_13), col = "purple")# + ylim(c(0, 25)) #+ xlim(c(-2, 5.2))
```

What's happening for small values of $x$ (and a bit for large values)? This is called [Runge's Phenomenon](https://en.wikipedia.org/wiki/Runge%27s_phenomenon) meaning that the boundary activity of high-order polynomials has very large derivatives. Let's go back to the same scale as before and see what's happening:

```{r}
basic + 
  coord_cartesian(xlim = c(-2, 5), ylim = c(-3, 25)) + 
  stat_function(fun = plot_function_degree_13, args = list(b = b_poly_13), col = "purple")
```

#Extrapolation vs Interpolation

Let's take a look at the Galton Data again.

```{r}
pacman::p_load(HistData, ggplot2)
data(Galton)
mod = lm(child ~ parent, Galton)
b_0 = mod$coefficients[1]
b_1 = mod$coefficients[2]
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Let's say I want to predict child's height for parents' average height of 70in. All I do is:

```{r}
predict(mod, data.frame(parent = 70))
```

What if I want to predict for a parents' height of 5in. Is there any 12in tall human being? No... it is absurd. But nothing stops you from doing:

```{r}
predict(mod, data.frame(parent = 5))
```

That's [actually possible](https://www.guinnessworldrecords.com/news/2012/2/shortest-man-world-record-its-official!-chandra-bahadur-dangi-is-smallest-adult-of-all-time/).

Look at our linear model from Euclid's perspective:

```{r}
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  xlim(-20, 120) + 
  ylim(-20, 120) +
  coord_equal(ratio = 1)
```

What is a linear model with $p + 1 = 2$. It's just a line. When geometry was first formalized by Euclid in the Elements, he defined a line to have "breadthless length" with a straight line being a line "which lies evenly with the points on itself". By "breadthless" he meant infinite in either direction. There is no mathematical problem with predicting childrens' heights using negative parents' heights e.g.

```{r}
predict(mod, data.frame(parent = -5))
```

But this is absurd. So now we need to talk about a fundamental concept in data science we've been kind of avoiding and one that most people ignore. There are two types of prediction: interpolation and extrapolation. Interpolation is essentially the type of "prediction" we've been talking about this whole class. 

Extrapolation is totally different. It's what happens when you predict outside of the range of the covariate data you've seen in $\mathbb{D}$. Extrapolation is very dangerous - your models only work based on $\mathbb{D}$. Extrapolation is prediction outside of the range you've seen before which means. You better have a good theoretical reason as to why your $\mathbb{H}$ function class will extend outside that range. Because each $\mathbb{H}$ function class will extrapolate very very differently.



What happens during extrapolation? Let's look at the (a) linear model, (b) polynomial model with degree 2 and (c) polynomial with degree 13.

```{r}
degree_2_poly_mod = lm(child ~ poly(parent, 2, raw = TRUE), Galton)
b_poly_2 = coef(degree_2_poly_mod)
degree_13_poly_mod = lm(child ~ poly(parent, 13, raw = TRUE), Galton)
b_poly_13 = coef(degree_13_poly_mod)
b_poly_13[is.na(b_poly_13)] = 0

plot_function_degree_2 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2
}
plot_function_degree_13 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4 + b[6] * x^5 + b[7] * x^6 + b[8] * x^7 + b[9] * x^8 + b[10] * x^9  + b[11] * x^10 + b[12] * x^11 + b[13] * x^12 + b[14] * x^13
}

# xymin = 65
# xymax = 71
xymin = 50
xymax = 90
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue") +
  coord_cartesian(xlim = c(xymin, xymax), ylim = c(xymin, xymax)) +
  stat_function(fun = plot_function_degree_2, args = list(b = b_poly_2), col = "red", xlim = c(xymin, xymax)) +
  stat_function(fun = plot_function_degree_13, args = list(b = b_poly_13), col = "orange", xlim = c(xymin, xymax))
```

Polynomial models have *TERRIBLE* extrapolation risk - totally unpredictable. Please do not use them if you think you'll ever extrapolate!!!

# Overfitting with Polynomials

Can we achieve $R^2 = 100\%$ using polynomial regression? Yes. Here's an example in one dimension. These are called "interpolation polynomials". In one dimension, as long as the $x$ values are distinct, $n$ data point can be fit by a $n - 1$ degree polynomial. Here's an example with a few data points:

```{r}
set.seed(1003)
n = 5
beta_0 = 1
beta_1 = 0
beta_2 = 1
x = runif(n)

h_star_x = beta_0 + beta_1 * x + beta_2 * x^2
y = h_star_x + rnorm(n)
#scatterplot it
df = data.frame(x = x, y = y, h_star_x = h_star_x)
basic = ggplot(df, aes(x, y)) +
  geom_point()
basic
```

Now fit polynomial models:

```{r}
degree_4_poly_mod = lm(y ~ poly(x, 5, raw = TRUE))
b_poly_4 = coef(degree_4_poly_mod)

plot_function_degree_4 = function(x, b){
  b[1] + b[2] * x + b[3] * x^2 + b[4] * x^3 + b[5] * x^4
}

basic + stat_function(fun = plot_function_degree_4, args = list(b = b_poly_4), col = "purple")
```

Perfect fit!

```{r}
summary(degree_4_poly_mod)$r.squared
```

This is the same thing we've seen before! If $n = p + 1$, then the design matrix is square and there is no need to project onto a lower dimensional subspace. To estimate the linear model, one only needs to solve $n$ equations with $n$ unknowns.

My recommendations:
1) Keep polynomial degree low. Preferably 2. Anything past 2 is not interpretable anyway. We didn't talk about "interpretability" of models yet, but you get the idea. If you are using degree = 2, use raw polynomials so you get interpretability (see next section).
2) Be very careful not to extrapolate: make sure future predictions have the measurements within range of the training data $\mathbb{D}$. Extrapolations are going to be very, very inaccurate. Polynomial regressions I'm sure have gotten data scientists fired before.



# Orthogonal vs raw polynomials

Why is orthonormal polynomial the default? You can argue that doing a QR decomposition on the polynomial expansion and employing Q in the design matrix will change b thereby making b uninterpretable! So why use orthonormal polynomials? Here's why:

```{r}
n = 1000
set.seed(1984)
X = data.frame(x = c(runif(n / 2, 0, 1e-2), runif(n / 2, 0, 1e6)))
d = 10
num_digits = 8
Xmm_orth = model.matrix(~ 0 + poly(x, d), X)
colnames(Xmm_orth)[1 : d] = 1 : d
Xmm_raw = model.matrix(~ 0 + poly(x, d, raw = TRUE), X)
colnames(Xmm_raw)[1 : d] = 1 : d
```

Let's look at the design matrix for small values of x:

```{r}
head(as.matrix(X))
round(head(Xmm_orth), num_digits)
round(head(Xmm_raw), num_digits)
```

You get numerical underflow almost immediately when using the raw polynomial computations (you get it by degree 4). And thus you can't even get the OLS estimates:


```{r}
y = rnorm(n)
solve(t(Xmm_raw) %*% Xmm_raw) %*% t(Xmm_raw) %*% y
```

Let's look at the design matrix for large values of x:

```{r}
tail(as.matrix(X))
round(tail(Xmm_orth), num_digits)
round(tail(Xmm_raw), num_digits)
```

You get numerical overflow in the design matrix (but it will happen later). But the second you start to use the design matrix with 10^59's inside...

```{r}
solve(t(Xmm_raw[800 : 1000, ]) %*% Xmm_raw[800 : 1000, ]) %*% t(Xmm_raw[800 : 1000, ]) %*% y
```

As opposed to

```{r}
solve(t(Xmm_orth) %*% Xmm_orth) %*% t(Xmm_orth) %*% y
```

No problem at all!!!

So that's the reason: numerical stability. But if you need interpretability, you need raw polynomials. But if you're interpreting the model, how do you even interpret beyond degree 2???

# Log transformations

We will be examining the diamonds dataset. Let's take a moment to read about our data. In real life, you will take more than just a moment to learn about the data. And more than just a moment to clean the data and do sanity checks. As these steps are more of the grunt work of data science (and are learned on the fly), I won't really cover them formally.

```{r}
rm(list = ls())
pacman::p_load(ggplot2) #this loads the diamonds data set too
?diamonds
dim(diamonds)
```

That's a huge $n$. So, let's expect things to take a bit longer when processing.

A natural increasing relationship will likely be found between weight and price. Let's see it visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()
```


How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ carat, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

What does the intercept say about extrapolation?

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") +ylim(0, 50000)
```

Maybe the relationship between weight and price is not linear - but exponential? E.g. if the weight of a diamond doubles, maybe the price quadruples? Or linear increases in weight yield percentage increases in price. This seems plausible. 

Besides the plausibility of the functional form, there is an agnostic reason to employ log y as the predictive target. Let's first examine the univariate data!


```{r}
skimr::skim(diamonds$price)
```

Very large standard error and very long tail which can be seed more clearly here:

```{r}
ggplot(diamonds) + geom_histogram(aes(price), binwidth = 200)
```

Let's take a look at the distribution after logging:

```{r}
ggplot(diamonds) + geom_histogram(aes(x = log(price)), binwidth = 0.01)
```

Some strange artifacts appear. Why the gap? Why is it "cut" sharply at a maximum? These are questions to ask the one who collected the data. But let's get back to the log story...

Popular wisdom says logging this type of highly skewed-right distribution would possibly make the model "more linear in x". Put another way, it would be easier to "catch" (predict) the long tail since it won't be a long tail anymore after you log-transform. It would also prevent observations with large y's becoming "leverage points" i.e. points that unduly influence the model and thereby warp its ability to predict the average observation. If we have time later in the semester, we can learn about leverage points, but not now. In Econ 382 you learn more reasons for why you should use log the response, but those reasons are relevant for inference so we won't discuss them here.

Let's give the model with ln(y) a whirl. Maybe we'll even learn something about diamonds. The way to create such a model is to simply fit an OLS model to log y. This is called a log-linear model. Since this is a pretty standard thing to do so R's formula notation has it built-in as follows:

```{r}
log_linear_mod = lm(log(price) ~ carat, diamonds)
b = coef(log_linear_mod)
b
```

Let's see what this looks like.

```{r}
ggplot(diamonds, aes(x = carat, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green") + ylim(6, 12)
```

It looks very bad if carat is large. That means our little theory about carats getting larger yielding multiples of price doesn't correspond to reality.

How did we do?

```{r}
summary(log_linear_mod)$r.squared
summary(log_linear_mod)$sigma
```

Look at that RMSE! That dropped like a rock! Is that real?

No. RMSE before is in the units of y. And now y is now in ln($). So this RMSE and the previous RMSE are *not* comparable.

The $R^2$ are *not* comparable either. Even though they're on a [0, 1] scale in both models, the SST's are different so you're measuring the proportion of a different variance.

Let's attempt to compare apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

What does this mean? It means this was a bad idea. Those residuals for large carats are insanely large. They're wrong on a log scale! Which means they're off by orders of magnitude. Working with logged y is dangerous business if you're wrong! Before you were off by a few thousand dollars; now you're off by millions. For example. Let's look at a large diamond:

```{r}
xstar = diamonds[diamonds$carat > 5, ][1, ]
xstar$price
predict(mod, xstar)
exp(predict(log_linear_mod, xstar))
```

That's a pretty bad residual!

How about log-log model? 

```{r}
log_log_linear_mod = lm(log(price) ~ log(carat), diamonds)
b = coef(log_log_linear_mod)
b
```

Let's see what it looks like:

```{r}
ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Well look at that! That's a nice looking model. (Note that the slope coefficients in log-log models, i.e. b_2 here, are called "elasticity" in Econ 382 as it measures how the relative change in x affects the relative change in y).

How are our metrics?

```{r}
summary(log_log_linear_mod)$r.squared
summary(log_log_linear_mod)$sigma
```

Let's see apples-to-apples to the natural y model.

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

This is on-par with the vanilla OLS model, but still doesn't "beat it". There was no guarantee that we would be "beat it" even though we used procedures that are reasonable and popular! My belief is that we really should be wary of the maximum price and employ that in the model. Maybe we'll do this in a lab exercise?

Let's repeat this entire exercise using the length of the diamond. The length of the diamond feature is confusingly named "x" in the dataset. It is an "x" but it's also the diamond's "x"!!!

```{r}
ggplot(diamonds, aes(x = x, y = price)) + 
  geom_point()
```

Besides the non-linear relationship, what else do you see? Mistakes in the dataset! Can a real diamond have zero length?? Yes. This is the real world. There are mistakes all the time.

Let's kill it! How many are we dealing with here?

```{r}
nrow(diamonds[diamonds$x == 0, ])
```


```{r}
diamonds = diamonds[diamonds$x != 0, ]
```

What's the deal with the x variable now?

```{r}
skimr::skim(diamonds$x)
```

How good does a best guess linear relationship do?

```{r}
mod = lm(price ~ x, diamonds)
b = coef(mod)
b
summary(mod)$r.squared
summary(mod)$sigma
```

Let's see the best fit line $g(x)$ visually:

```{r}
ggplot(diamonds, aes(x = x, y = price)) + geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

Again we got some bad extrapolation going on which we can't fix using a purely linear modeling strategy.

Let's log-linearize it and see how we do.

```{r}
log_linear_mod = lm(log(price) ~ x, diamonds)
b = coef(log_linear_mod)
ggplot(diamonds, aes(x = x, y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? Ensure it's apples-apples.

```{r}
log_y_hat = log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

Still not better. Log-log?

```{r}
log_log_linear_mod = lm(log(price) ~ log(x), diamonds)
b = coef(log_log_linear_mod)
ggplot(diamonds, aes(x = log(x), y = log(price))) + 
  geom_point() + 
  geom_abline(intercept = b[1], slope = b[2], col = "green")
```

How did we do? 

```{r}
log_y_hat = log_log_linear_mod$fitted.values
y_hat = exp(log_y_hat)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
```

We did it. We found a log transformation that seems to give higher predictive power than the vanilla linear model on the raw repsonse and raw feature.

This brings up the whole idea of "model selection". We went hunting for models until we found one that's better. We will hopefully do model selection today...

Transforming y is a big decision as it changes the response metric! The rule of thumb is it is easier to model a response metric that has less extreme values (especially when using linear models) as the extreme values have a big impact on slope coefficients and can distort the best fit line due to the least squares minimization (hence the popularity of logging the response).

Let's see if we get anywhere with this using all the features in this model.

```{r}
lm_y = lm(price ~ ., diamonds)
lm_ln_y = lm(log(price) ~ ., diamonds)
summary(lm_y)$r.squared
summary(lm_y)$sigma

#now for the log-linea model
y_hat = exp(lm_ln_y$fitted.values)
e = diamonds$price - y_hat
SSE = sum(e^2)
SST = sum((diamonds$price - mean(diamonds$price))^2)
Rsq = 1 - sum(e^2) / SST
Rsq
RMSE = sqrt(SSE / (nrow(diamonds) - 2))
RMSE
``` 

This is pretty convincing evidence that this transformation does a better job (at least in our linear modeling context).

Let's look at one prediction:

```{r}
predict(lm_y, diamonds[12345, ])
exp(predict(lm_ln_y, diamonds[12345, ]))
diamonds$price[12345]
```

Again, we should be careful when you use $g$ after logging, you will have to exponentiate the result (middle line above). 

Small point: this exponentiation is known to create bias because $E[Y]$ is different from $exp(E[ln(y)])$ (for those who took 368 - remember Jensen's inequality?) For the purposes of this class, this can be ignored since we are evaluating g on its own merits and we're doing so honestly.

If you like this stuff, there are a whole bunch of transformations out there that are even cooler than the natural log. 

# Linear Models with Feature Interactions

Let's go back to modeling price with weight. Let us add a third variable to this plot, color, a metric about the "yellowness" of the diamond. This is an ordinal categorical variable ranging from D (most clear i.e. best) to J (most yellow in this dataset i.e. worst).


```{r}
pacman::p_load(ggplot2)
base = ggplot(diamonds, aes(x = carat, y = price)) 
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div")
```

We can split the data on color to see it more clearly:

```{r}
base +
  geom_point() +
  facet_wrap(~ color, ncol = 3) + 
  aes(color = color) + scale_color_brewer(type = "div")
```


What do we see here? It looks like the slope of the price vs. carat linear model is slightly affected by color. For instance, the "D" color diamonds' price increases much faster as weight increases than the "E" color diamonds' price increases in weight, etc. Why do you think this is?

We can picture two of these linear models below by fitting two submodels, one for D and one for J:

```{r}
mod_D = lm(price ~ carat, diamonds[diamonds$color == "D", ])
b_D = coef(mod_D)
mod_J = lm(price ~ carat, diamonds[diamonds$color == "J", ])
b_J = coef(mod_J)
b_D
b_J
```

Let's see it on the plot:

```{r}
base +
  geom_point(aes(col = color)) + scale_color_brewer(type = "div") +
  geom_abline(intercept = b_D[1], slope = b_D[2], col = "blue", lwd = 2) +
  geom_abline(intercept = b_J[1], slope = b_J[2], col = "red", lwd = 2)
```

This indicates a separate intercept and carat-slope for each color. How is this done? Interacting carat and slope. The formula notation has the `*` operator for this. It is multiplication in formula land after all!

```{r}
mod = lm(price ~ color, diamonds)
coef(mod)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #beware: strange naming convention on the interaction terms for an ordered factor
diamonds$color = factor(diamonds$color, ordered = FALSE)
mod = lm(price ~ carat * color, diamonds)
coef(mod) #much better...
```

The reference category is color D. This means every other color should start lower and have a lower slope. This is about what we see above.

How much of a better model is this than a straight linear model?

```{r}
mod_vanilla = lm(price ~ carat + color, diamonds)
summary(mod_vanilla)$r.squared
summary(mod_vanilla)$sigma
summary(mod)$r.squared
summary(mod)$sigma
```

You can get more predictive accuracy out of this. We added a degree of freedom? Is this gain real? Yes. With one more feature and $n = 54,000$ there is no chance this gain came from overfitting noise. Add 10,000 garbage features, yes, there will be overgitting.

Let's take a look at carat with another variable, depth, a continuous predictor. High depth indicates diamonds are skinny and tall; low depth indicates diamonds are flat like a pancake.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = depth), lwd = 1, alpha = 0.5) + scale_colour_gradientn(colours = rainbow(5))
```

It seems people like flatter diamonds and are willing to pay more per carat. Let's see this in the regression:

```{r}
mod = lm(price ~ carat + depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * depth, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

If carat increases by one unit, how much does price increase by? A tiny amount of increase.

How about cut?


```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(col = cut), lwd = 0.5) + scale_color_brewer(type = "div")
```

Likely something here.

```{r}
mod = lm(price ~ carat, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Yes.

Can we include all these interactions?

```{r}
mod = lm(price ~ carat + color + depth + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ carat * (color + depth + cut), diamonds)
summary(mod)$r.squared
summary(mod)$sigma
coef(mod)
```

A decent gain once again.

What does the design matrix look like there? What is $p$?

```{r}
diamonds$cut = factor(diamonds$cut, ordered = FALSE)
Xmm = model.matrix(price ~ carat * (color + depth + cut), diamonds)
head(Xmm)
```


Can we take a look at interactions of two categorical variables?


```{r}
plot1 = ggplot(diamonds, aes(x = cut, y = color)) +
  geom_jitter(aes(col = price), lwd = 0.5) + scale_colour_gradientn(colours = rainbow(5))
plot1
```

Cool animation possible. May not work because it needs a ton of packages...

```{r}
pacman:::p_load_gh("dgrtwo/gganimate")
plot1 + transition_time(price)
```

Not so clear what's going on here. Let's see what the regressions say:


```{r}
mod = lm(price ~ color + cut, diamonds)
summary(mod)$r.squared
summary(mod)$sigma
mod = lm(price ~ color * cut, diamonds)
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

Not too much gain.
