---
title: "Practice Lectures Week 3 MATH 342W Queens College"
author: "Professor Adam Kapelner"
date: "Feb 14, 2022"
---




## Errors and Warnings

You can write better functions if you make use of errors and warnings. Java forces you to catch likely errors via the "throws" designation for a method but there is no such requirement in R.

* Errors are unrecoverable, they halt execution i.e. red lights
* Warnings (under usual execution) do not halt execution, but they display a message, i.e. yellow lights

Here's how they work:

```{r}
my_vector_sum = function(xs){
  
  if (!(class(xs) %in% c("numeric", "integer"))){ #short for class(xs) == "numeric" | class(xs) == "integer"
    stop("You need to pass in a vector of numbers not a vector of type \"", class(xs), "\".\n") #throw error!
    # warning("Your vector of type \"", class(xs), "\" will be coerced to numbers.\n") #throw warning!
    # xs = as.numeric(as.factor(xs))
  }
  
  tot = 0
  for (x in xs){
    tot = tot + x
  }
  tot
}
my_vector_sum(c(1, 2, 3))
my_vector_sum(c("a", "b", "c"))
```

There is a try-catch as well:

```{r}
xs = c("a", "b", "c")
tot = my_vector_sum(xs)

tot = tryCatch(
  {
    my_vector_sum(xs)
  },
  error = function(e){
    print("xs was non-numeric... coercing xs to numeric...")
    my_vector_sum(as.numeric(as.factor(xs)))
  }
)
tot
```

The recommended thing to do of course is to query if it is non-numeric within the function `my_vector_sum` and cast it then. Possibly create an argument toggling this behavior on/off with a default of off.

## Matrix operations in R

R can do all the standard matrix operations. Let's go through them quickly. First initialize two example matrices:

```{r}
A = matrix(rep(1, 4), nrow = 2)
A
B = array(seq(1, 4), dim = c(2, 2))
B
I = diag(2) #create an identity matrix of size 2x2
I
```

Now we show off some operations:

```{r}
A * B #element-wise multiplication
A %*% B #matrix multiplication
B %*% I
t(B) #transpose
solve(B)
solve(A) #BOOM - why?
tryCatch(
  {
    solve(A)
  },
  error = function(e){
    print("matrix not invertible, doing Moore-Penrose generalized pseudoinverse instead...")
    MASS::ginv(A)
  }
)


#how would you wrap this and handle it in the real world?
solve(I)
#rank(A) = 1 #no such function... but... there are tons of add-on libraries for matrix computations e.g.
pacman::p_load(Matrix) #load the Matrix library
rankMatrix(B)
rankMatrix(A)
rankMatrix(I)
```

Note that vectors and matrices are not the same:

```{r}
v = c(1, 2, 3) #3-d vector
t(v) #converts to 1x3 vector... unsure why
t(t(v))
v %*% v #seems to default to dot product
t(v) %*% t(t(v)) #dot product
I = diag(3)
I %*% v #seems to default correctly!
I %*% t(v) #actually uncomformable
```


## Support Vector Machines (SVM)

You will code a basic SVM for homework. Here we use the `e1071` library.

```{r}
pacman::p_load(e1071)
```

We make a simple dataset first.

```{r}
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
```

We haven't spoken about visualization yet, but it is important we do some of it now. First we load the visualization library we're going to use:

```{r}
pacman::p_load(ggplot2)
```

And let's plot the data:

```{r}
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

Now we fit a linear SVM. Since it is linearly separable, we can make $\lambda = 0$. Since the package doesn't allow zero, we make it trivially small. 

```{r}
Xy_simple_feature_matrix = as.matrix(Xy_simple[, 2 : 3])
lambda = 1e-9
n = nrow(Xy_simple_feature_matrix)
svm_model = svm(Xy_simple_feature_matrix, Xy_simple$response, kernel = "linear", cost = (2 * n * lambda)^-1, scale = FALSE)
```

The model object can be queried to find the "support vectors" i.e. the observations that lie on the wedge. Let's visualize them too.

```{r}
svm_model$index
Xy_simple$is_support_vector = rep("no", n)
Xy_simple$is_support_vector[svm_model$index] = "yes"
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response, shape = is_support_vector)) + 
  geom_point(size = 5)
simple_viz_obj
```


Now we calculate the weight vector. This is technical and not covered in the class yet (or maybe never):

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% Xy_simple_feature_matrix[svm_model$index, ] # the other terms
)
w_vec_simple_svm
w_norm = sqrt(sum(w_vec_simple_svm^2))
w_norm
```

We can also plot it. We have to convert from Hesse Normal form back into point-intercept form. Note that $b$ is the first entry of the `w_vec_simple_svm` vector

```{r}
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")
simple_viz_obj + simple_svm_line
```

We can also plot the wedge by plotting the top line (where b is augmented by 1) and the bottom line (where b is diminished by 1).

```{r}
simple_svm_top_line = geom_abline(
    intercept = -(w_vec_simple_svm[1] + 1) / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "yellow")
simple_svm_bottom_line = geom_abline(
    intercept = -(w_vec_simple_svm[1] - 1) / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "yellow")
simple_viz_obj + simple_svm_line + simple_svm_top_line + simple_svm_bottom_line
```



To understand the hyperparameter, let's introduce another data point so the training data is no longer linearly separable.

```{r}
Xy_simple = rbind(Xy_simple, c(0, 3.2, 3.2))
```

and plot it:

```{r}
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

Let's try SVM at different $\lambda$ values.

```{r}
lambda = 0.001#0.001 #1.3

Xy_simple_feature_matrix = as.matrix(Xy_simple[, 2 : 3])
n = nrow(Xy_simple_feature_matrix)
svm_model = svm(Xy_simple_feature_matrix, Xy_simple$response, kernel = "linear", cost = (2 * n * lambda)^-1, scale = FALSE)
summary(svm_model)
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% Xy_simple_feature_matrix[svm_model$index, ] # the other terms
)
sqrt(sum(w_vec_simple_svm^2))

Xy_simple$is_support_vector = rep("no", n)
Xy_simple$is_support_vector[svm_model$index] = "yes"
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response, shape = is_support_vector)) + 
  geom_point(size = 5)

simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")
simple_svm_top_line = geom_abline(
    intercept = -(w_vec_simple_svm[1] + 1) / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "yellow")
simple_svm_bottom_line = geom_abline(
    intercept = -(w_vec_simple_svm[1] - 1) / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "yellow")
simple_viz_obj + simple_svm_line + simple_svm_top_line + simple_svm_bottom_line
```

What lesson did we learn here? This hyperparameter really matters! (In the method's documentation note, it says "Parameters of SVM-models usually must be tuned to yield sensible results!") And it is not clear to me what the "wedge" is anymore when there isn't separability, nor what the support vectors are, and nor how `cost` parameter has much to do with the lambda from the notes. What is clear is that we need to figure out a way to deal with selecting the "right" hyperparameter value automatically. So far neither the perceptron nor the SVM is an algorithm for binary classification that comes without flaws.


## Nearest Neighbor algorithm

Load up the breast cancer data set again.

```{r}
Xy = na.omit(MASS::biopsy) #The "breast cancer" data with all observations with missing values dropped
X = Xy[, 2 : 10] #V1, V2, ..., V9
y_binary = as.numeric(Xy$class == "malignant")
```

Let's say we want to build a nearest neighbor model with the first covariate only. We are then looking for the label (response) of the closest x_1. Here is a simple function that does it:

```{r}
nn_function = function(x_star){
  y_binary[which.min((X[, 1] - x_star)^2)]
}
nn_function(7.8)
nn_function(5.2)
```

Why is this silly for this dataset?

```{r}
str(X)
```

The features are not truly continuous. Would it make sense in higher dimensions? Your homework...

Has this been coded before? Definitely...

```{r}
pacman::p_load(class)
?knn
```

We can fit a knn model *and* predict in one shot via:

```{r}
y_hat = knn(X, c(4, 2, 1, 1, 2, 1, 2, 1, 1), y_binary, k = 1)
y_hat
```

Why is build model and predict in one shot natural in knn?

Now for an interesting exercise that will setup future classes:

```{r}
y_hat = knn(X, X, y_binary, k = 1)
y_hat
all.equal(y_hat, factor(y_binary))
```

No errors! Can this be a good model? No... "something" must be wrong! It is too good to be true.

Something is wrong. This is the first example of "overfitting". We will explore this later in depth (it is one of the core concepts of this course).

Let's see $K > 1$


```{r}
y_hat = knn(X, X, y_binary, k = 10)
y_hat
all.equal(y_hat, factor(y_binary))
```

Why would there be difference now between predictions and the actual data?

```{r}
rm(list = ls())
```



## Simple Linear Regression (p = 1)

To understand what the algorithm is doing - best linear fit by minimizing the squared errors, we can draw a picture. First let's make up some very simple training data $\mathbb{D}$.

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
h_star = beta_0 + beta_1 * x
epsilons = rnorm(n, mean = 0, sd = 0.33)
y = h_star + epsilons 
```

And let's plot the data:


```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

And its true $h^*$ line:

```{r}
true_hstar_line = geom_abline(intercept = beta_0, slope = beta_1, color = "green")
simple_viz_obj + true_hstar_line
```

Now let's calculate the simple least squares coefficients:

```{r}
r = cor(x, y)
s_x = sd(x)
s_y = sd(y)
ybar = mean(y)
xbar = mean(x)

b_1 = r * s_y / s_x
b_0 = ybar - b_1 * xbar
b_0
b_1
```

Note how $b_0$ and $b_1$ are not exactly the same as $\beta_0$ and $\beta_1$. Why?

And we can plot it:


```{r}
simple_ls_regression_line = geom_abline(intercept = b_0, slope = b_1, color = "red")
simple_viz_obj + simple_ls_regression_line + true_hstar_line
```

Review of the modeling framework:

The difference between the green line and red line is the "estimation error". The difference between the green line and the points is a combination of error due to ignorance and error due to misspecification of $f$ as a straight line. In most real-world applications, estimation error is usually small relative to the other two. In the era of "big data", $n$ is usually big so estimation error is pretty small.

Recall that the noise (epsilons) are the difference between the data and the green line:

```{r}
simple_df$hstar = beta_0 + beta_1 * simple_df$x
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
epsilon_line_segments = geom_segment(aes(xend = x, yend = hstar), position = position_nudge(x = 0.002))
simple_viz_obj + epsilon_line_segments + true_hstar_line
```

And that the residuals (e's) are the difference between the measurements of the response in the actual data and the green line:

```{r}
simple_df$gs = b_0 + b_1 * simple_df$x
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
e_line_segments = geom_segment(aes(xend = x, yend = gs), color = "purple")
simple_viz_obj + simple_ls_regression_line + e_line_segments
```

Examining both at the same time:

```{r}
simple_viz_obj + simple_ls_regression_line + true_hstar_line + e_line_segments + epsilon_line_segments
```


## Assessing quality of a simple linear regression

Note: our $\mathcal{A}$ was ordinary least squares. What follows below is a method of assessing model fit quality not only for the least squares line, or any linear fit, but any regression fit.

```{r}
simple_df$gs = b_0 + b_1 * simple_df$x
simple_df$e = y - simple_df$gs

sse = sum(simple_df$e^2)
mse = sse / (n - 2)
rmse = sqrt(mse)
sse
mse
rmse

s_sq_y = var(y)
s_sq_e = var(simple_df$e)
rsq = (s_sq_y - s_sq_e) / s_sq_y
rsq

#calculated in a different, but equivalent way
sse_0 = (n - 1) * s_sq_y
(sse_0 - sse) / sse_0
```

Let's take a look at $R^2$ visually. We compute null residuals (the $e_0$'s) and model residuals (the $e$'s) and plot a them.

```{r}
simple_df$e_0 = y - mean(y)

ggplot(simple_df) + 
  geom_histogram(aes(x = e), fill = "darkgreen", alpha = 0.3) + 
  geom_histogram(aes(x = e_0, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")

ggplot(simple_df) + 
  stat_density(aes(x = e), fill = "darkgreen", alpha = 0.3) + 
  stat_density(aes(x = e_0, fill = "red", alpha = 0.3)) +
  theme(legend.position = "none")
```

Note residuals always have sample average = 0 (modulo numeric error):

```{r}
mean(simple_df$e_0)
mean(simple_df$e)
```

We will prove this fact later in class.

Let's take a look at predictions by truth:


```{r}
ggplot(simple_df, aes(x = gs, y = y)) + 
  geom_point() + 
  xlim(0, max(simple_df$gs, y)) + 
  ylim(0, max(simple_df$gs, y)) + 
  xlab("yhat") + 
  coord_fixed() + 
  geom_abline(intercept = 0, slope = 1, color = "orange")
```

Linear regression is pretty popular so there's obviously support for this in R. Before we talk about this, we need to discuss another object type in R. It is called the "formula" object. Here's an example:

```{r}
simple_model_formula = as.formula("y ~ x")
simple_model_formula
```

You can use a convenience:

```{r}
simple_model_formula = y ~ x
simple_model_formula
```

How did this work? R interprets this as a formula because it sees the tilde inside (you have to dig pretty deep into the R language to understand the tilde operator but feel free to ignore it now).

By default the formula object when executed prints out the string you supplied. But obviously it is not just a string. This object contains instructions to model `y` with `x`. This may seem opaque now but you'll see how this works soon.

Getting back to support for the default linear model. The popular function for that implements least squares linear regression is loaded into R automatically in the package `stats`. Here is a list of packages that are loaded into R by default (save the `ggplot2` and RStudio's addition):

```{r}
search()
```

The function `lm` runs least squares. Let's see it:

```{r}
simple_linear_model = lm(simple_model_formula)
simple_linear_model
```

You can skip a step by putting the formula in-line (this is how it's usually done):


```{r}
simple_linear_model = lm(y ~ x)
simple_linear_model
class(simple_linear_model)
```

What did this do? By specifying the formula that $y$ should be modeled with $x$ sing the simple linear model $y = w_0 + w_1 x$
By default it prints out $b_0$ and $b_1$. You can store the vector via:

```{r}
b = coef(simple_linear_model)
b
names(b)
class(b) #i.e. a vector of numbers dimension 2 where each entry is named
```

You can query the linear model about its fit as well:

```{r}
names(summary(simple_linear_model))
summary(simple_linear_model)$r.squared #the R^2
summary(simple_linear_model)$sigma #the RMSE
```

Cleanup...

```{r}
rm(list = ls())
```


## Simple Linear Regression with an example data set

Load up the famous Boston Housing data

```{r}
?MASS::Boston
Xy = MASS::Boston
head(Xy)
```

We would like to see how each feature relates to the response, `medv`. This is a quick and dirty way to do it:

```{r}
for (feature in setdiff(colnames(Xy), "medv")){
  plot(ggplot(Xy, aes(x = Xy[, feature], y = medv)) + geom_point() + xlab(feature))
}
```

Let's try to explain `medv` using the feature `rm` in a simple linear regression (least squares) model.

```{r}
x = Xy$rm
y = Xy$medv

r = cor(x, y)
s_x = sd(x)
s_y = sd(y)
ybar = mean(y)
xbar = mean(x)

b_1 = r * s_y / s_x
b_0 = ybar - b_1 * xbar
b_0
b_1
```

and we can plot this line atop the data:

```{r}
simple_viz_obj = ggplot(Xy, aes(x = rm, y = medv)) + geom_point()
simple_ls_regression_line = geom_abline(intercept = b_0, slope = b_1, color = "red")
simple_viz_obj + simple_ls_regression_line
```

And how well did we do?

```{r}
yhat = b_0 + b_1 * x #this is the g(x^*) function!
e = y - yhat
sse = sum(e^2)
mse = sse / length(y)
rmse = sqrt(mse)
sse
mse
rmse

s_sq_y = var(y)
s_sq_e = var(e)
rsq = (s_sq_y - s_sq_e) / s_sq_y
rsq
```

SSE is not a super useful number alone. MSE is not super useful alone. RMSE is... what does it mean? What does $R^2$ mean?


```{r}
Xy$null_residuals = y - mean(y)
Xy$residuals = e

ggplot(Xy) + 
  stat_density(aes(x = residuals), fill = "darkgreen", alpha = 0.6, adjust = 0.5) + 
  stat_density(aes(x = null_residuals, fill = "red", alpha = 0.6, adjust = 0.5)) +
  theme(legend.position = "none")
```

This is not a great model. Why? Three sources of error... what do you think are the biggest sources of error?

```{r}
rm(list = ls())
```

Let's do this again using R's `lm` function. Here we can leverage the data frame object by using the names of the variables in our model formula:

```{r}
mod = lm(medv ~ rm, data = MASS::Boston)
```

I read this as "build a linear model where we explain median household value using the average number of rooms in the Boston housing dataset". One line! We can of course ping the model for everything we've been talking about:

```{r}
coef(mod)
summary(mod)$r.squared
summary(mod)$sigma
```

And ggplot has amazing integration. Here it runs the model internally and gives smoothed confidence bands (we did not discuss this):

```{r}
ggplot(MASS::Boston, aes(rm, medv)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

```{r}
rm(list = ls())
```







